Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-09-21 main/MSstatsPTM_Main.tex
Minor updates to main manuscript

%\author{
%  LastName1, FirstName1\\
%  \texttt{first1.last1@xxxxx.com}
%  \and
%  LastName2, FirstName2\\
%  \texttt{first2.last2@xxxxx.com}
%}
\date{}
\begin{document}

@@

The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was then compared against the commonly applied t-test. The results demonstrate that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs. The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.

@@

\todo{Not sure what to do with this part and figure. How would this look in a paper?}
Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \ref{fig:data-structure}. Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference. The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts. The standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is

\[
\left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]

@@

The proposed statistical approach methods were evaluated using computer simulation. Specifically, their properties under adjustment with respect to protein abundance and batch effects were evaluated.

@@

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary \todo{link to Sec. S4.1}.

In this simulation, when there was no change in protein abundance, all the considered approaches well calibrated the Type I error rate \todo{figref Fig. 4a}. However, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig. 4b)}. As shown in \todo{figref Fig. 4c-e}, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data \todo{figref (Fig. 4d-e)}. In cases with small sample sizes (e.g., 2 replicates) as shown in \todo{figref Fig. 4d}, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in \todo{figref Fig. 4e} [with 4 conditions] over \todo{figref Fig. 4d} [with 2 conditions]). Further results are provided in Supplementary \todo{link to Sec. S4.1}.

\paragraph*{Computer simulation: batch effects.}

PTM experiments often create batches during data acquisition. The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data. Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.\todo{Find source for batch analysis} Although simple, these ad-hoc methods lack statistical justification. We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec. S4.2}. Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).

The major results are summarized in \todo{figref Fig. 5} with additional details provided in Supplementary \todo{link to Sec. S4.2}. In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig. 5a). In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig. 5b). As shown Fig. 5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability. Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig. \todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs. S8-S9}). The negative impact was only partially reduced by increasing the sample size to 5. The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.

>>>>

\author{
 Devon Kohler\\
 \and
 Tsung-Heng Tsai\\
 \and
 Ting Huang\\
 \and
 Erik Verschueren
 \and
 Trent Hinkle
 \and
  Meena Choi\\
 \and
 Olga Vitek
}
\date{}
\begin{document}

@@

The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was then compared against the commonly applied t-test and limma package. The results demonstrate that by appropriately leveraging the information from the entire dataset the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes, results in a better calibrated type I error rate, and improves the statistical power of detecting changes in PTMs. The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.

@@

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired. If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \ref{fig:data-structure}. Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference. The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.

@@

The proposed statistical approach methods were evaluated using computer simulation. Specifically, their properties under adjustment with respect to protein abundance and batch effects were evaluated. Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.

@@

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverages measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Details of the simulation can be found in Supplementary %\cref{supp-sec:sim}.

In this simulation, when the modification changes were entirely due to the changes in protein abundance across conditions, analysis without accounting for the protein-level changes resulted in off-target, high false positive rates \todo{figref (Fig. 4b)}. As shown in \todo{figref Fig. 4c-e}, in detection of systematic changes in PTM abundance, the proposed approach improved statistical power with small sample sizes in almost all the considered scenarios. The advantage by using the proposed approach over t-test became more profound in the presence of missing data \todo{figref (Fig. 4d-e)}. In cases with small sample sizes (e.g., 2 replicates) as shown in \todo{figref Fig. 4d}, performance by t-test decreased dramatically with one missing value, and increasing the log-fold change to 2 did not effectively reduce the negative impact. Two-sample t-test only used data within the groups of interest while ignoring the rest of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions (see for example, the improved performance in \todo{figref Fig. 4e} [with 4 conditions] over \todo{figref Fig. 4d} [with 2 conditions]). Further results are provided in Supplementary \todo{link to Sec. S4.1}.

%\paragraph*{Computer simulation: batch effects.}

%PTM experiments often create batches during data acquisition. The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data. Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.\todo{Find source for batch analysis} Although simple, these ad-hoc methods lack statistical justification. We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec. S4.2}. Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).

%The major results are summarized in \todo{figref Fig. 5} with additional details provided in Supplementary \todo{link to Sec. S4.2}. In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig. 5a). In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig. 5b). As shown Fig. 5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability. Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig. \todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs. S8-S9}). The negative impact was only partially reduced by increasing the sample size to 5. The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.
